---
title: "assignment-2a"
output: html_document
date: "2026-09-10"
author: "Kenny Jeannite"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DBI)
library(RPostgres)
library(dplyr)
con <- dbConnect(RPostgres::Postgres(), 
                 dbname = "school", 
                 host = "localhost", 
                port = 5432, 
                user = "postgres", password = rstudioapi::askForPassword())
```

# Introduction

Not every movie appeals to every viewer. By asking respondents to rate each movie on a scale from 1 (terrible) to 5 (amazing), I can identify which of six recent releases are best received within my sample. I plan to collect responses through a Google Form using a radio-button scale, with an additional option for "did not watch this movie." The core question I want to answer is: among the people I surveyed, which of these six recent releases is the most well-liked, and how much does that opinion vary from person to person?

One data challenge I anticipate is that most respondents will not have seen every one of the six movies, resulting in incomplete rating data for each person. To handle missing ratings, I will store unseen movies as NULL in the SQL database rather than as 0, since 0 would misrepresent "no opinion" as "worst possible rating" and skew average scores downward. When loading the data into R, these will appear as NA, and I will use na.rm = TRUE when computing averages so that unseen movies don't distort the results. I will also report the number of respondents who rated each movie, since an average based on a small number of ratings carries less weight than one based on a large number.


# Data Collection

I collected the data using a Google Form that I shared with a group of friends. Each respondent was asked to rate six recent movies on a scale from 1 to 5, where 1 represented a very negative opinion and 5 represented a very positive opinion. Respondents also had the option to indicate that they had not seen a movie rather than being forced to provide a rating.

In total, I collected 10 responses. Because not every respondent had seen every movie, some movies received fewer ratings than others. I treated these unanswered movie ratings as missing data rather than assigning them a score, since not seeing a movie does not reflect a positive or negative opinion of it.

## Table Creation and Loading the Data 

Below I am creating the three tables users, movies, and ratings, The survey was anonymized on Google forms so I did not get any of the individuals names. I exported the data via .csv and manually put the data into the database.

```{sql connection=con, eval=FALSE}
CREATE TABLE users (
    user_id SERIAL PRIMARY KEY,
    response_timestamp TEXT
);
```

```{sql connection=con, eval=FALSE}
CREATE TABLE movies (
    movie_id SERIAL PRIMARY KEY,
    title TEXT NOT NULL UNIQUE
);
```

```{sql connection=con, eval=FALSE}
CREATE TABLE ratings (
    user_id  INT REFERENCES users(user_id),
    movie_id INT REFERENCES movies(movie_id),
    rating   SMALLINT CHECK (rating BETWEEN 1 AND 5),
    PRIMARY KEY (user_id, movie_id)
);
```

I manually inserted the user_ids, movie_ids, and ratings here from the survey responses. 

```{sql connection=con, eval=FALSE}
INSERT INTO ratings (user_id, movie_id, rating) VALUES
(1,1,5),(1,2,4),(1,4,5),(1,5,4),(1,6,5),
(2,1,5),(2,3,3),(2,4,4),(2,6,5),
(3,1,5),(3,2,5),(3,4,3),(3,6,5),
(4,1,4),(4,2,5),(4,5,4),(4,6,5),
(5,1,3),(5,2,3),(5,5,1),(5,6,5),
(6,1,4),(6,2,3),(6,4,3),(6,6,5),
(7,5,4),(7,6,5),
(8,1,4),(8,2,5),(8,6,5),
(9,1,3),(9,2,3),(9,6,4),
(10,1,5),(10,2,5),(10,4,5),(10,5,3),(10,6,5);
```


### Loading the data 

I used a CROSS JOIN here to take every user and pair them with every movie in the survey. Since there were 10 respondents and 6 movies, this creates all 60 possible user-movie combinations. I then used a LEFT JOIN to bring in the actual rating when one existed. If a respondent did not rate a movie, there is no matching row in the ratings table, so the rating appears as NA when the data is loaded into R. This allows me to keep track of missing ratings without treating an unseen movie as a score of 0.

``` {r}
movie_ratings <- dbGetQuery(con, "
SELECT
    u.user_id,
    m.title,
    r.rating
FROM users u
CROSS JOIN movies m
LEFT JOIN ratings r
    ON u.user_id = r.user_id
    AND m.movie_id = r.movie_id
ORDER BY u.user_id, m.movie_id;
")

movie_ratings

```

#### Movie Summary 

To compare the six movies, I calculated the average rating, numbers of responses, standard deviation, and missing ratings. With the chart below you can tell that Obsession was a clear standout with an average rating of 4.9 and The Invite was the least liked.  Backrooms received the highest standard deviation among hte movies with multiple responses at 1.30. This indicated that it is widely varied between the respondents. 

``` {r}
movie_summary <- movie_ratings %>%
  group_by(title) %>%
  summarise(
    number_of_ratings = sum(!is.na(rating)),
    average_rating = mean(rating, na.rm = TRUE),
    standard_deviation = sd(rating, na.rm = TRUE),
    missing_ratings = sum(is.na(rating))
  ) %>%
  arrange(desc(average_rating))

movie_summary

```


# Conclusion 

This was an interesting and enjoyable assignment because it allowed me to use data collected from my friends to see which movies they liked, disliked, or had not seen. By looking at the average ratings, number of responses, and variation in the ratings, I was able to compare how each movie was received within my sample.

One thing I struggled with during the assignment was understanding why my SQL code blocks kept throwing errors in the R Markdown document. I eventually realized that I needed to add connection=con to the SQL code chunk headers so that each query could connect to the PostgreSQL database. Once I fixed that, I was able to run the SQL queries directly in the document and use the results in R for my analysis.


